Efficient Detection and Classification of Epigenomic Changes Under Multiple Conditions

### 2021-03-05 Latex/paper/paper.tex
txt update

% Paragraph 3: Introduce ChIP-seq
To quantify local epigenomic activity, a common high-throughput assay is chromatin immunoprecipitation followed by massively parallel sequencing (ChIP-seq). ChIP-seq begins with cross-linking DNA and proteins within chromatin structures, which are then fragmented by sonication. DNA fragments bound to the protein of interest are isolated by immunoprecipitation and sequenced via high-throughput sequencing to generate short reads pertaining to the original fragments. Sequences are then mapped onto a reference genome to determine their likely locations of origin. Genomic coordinates containing a high density of mapped reads (enrichment regions, peaks) indicate likely locations of protein-DNA interaction sites, while other regions are referred to as background regions. This read density is often summarized by counting reads mapped onto non-overlapping windows of fixed length (window read counts), forming the basis for downstream analyses. Across multiple conditions, regions exhibiting enrichment in at least one condition, but not across all conditions, indicate the presence of differential activity pertaining to the protein-DNA interaction of interest.

@@

% Paragraph 4: What we propose? Broad peaks, mixture model, and multiple conditions
Here, we present epigraHMM (Figure~\ref{fig:fig_motivation}), an efficient and flexible statistical method to identify differential regions of enrichment from epigenomic experiments with diverse signal profiles. We overcome the limitations of current DPCs with three major features. First, it uses a hidden Markov model (HMM) to account for the diversity in differential enrichment from broad and short ChIP-seq data collected under multi-replicate, multi-condition settings. Second, it models specific combinatorial patterns of enrichment via a finite mixture model emission distribution within the HMM differential state. Each mixture component pertains to a particular differential combinatorial pattern originated by the presence or absence of local enrichment across conditions, where a generalized linear model (GLM) encodes specific differential combinatorial patterns while accounting for sample- and window-specific normalizing offsets. Third, it simultaneously detects and classifies epigenomic changes under three or more conditions, a novelty not yet available in any other DPC algorithm. epigraHMM offers additional benefits over current HMM algorithms \citep{song2011identifying,allhoff2016differential} with an embedded GLM mixture model that allows the modeling of covariates of interest, the inclusion of normalizing offsets for non-linear biases (such as GC-content bias), and a fast parameter estimation scheme via rejection-controlled EM algorithm (RCEM).

@@

Using ChIP-seq data pertaining to histone modifications H3K27me3 and H3K36me3, and to EZH2 from the ENCODE Consortium \citep{landt2012chip}, we find that current DPCs have difficulty in accurately detecting broad regions of differential enrichment between several common cell lines (Figure~\ref{fig:fig_motivation}). In line with previous findings \citep{steinhauser2016comprehensive}, we observe that even current DPCs designed for broad data \citep{song2011identifying,allhoff2016differential} tend to detect either overly fragmented differential peaks or call regions exhibiting no difference in experimental signal between conditions as differential (Sections \ref{sec:sim2} and \ref{sec:broad}). Methods that rely on candidate peaks may also exhibit a compromised performance due to the limitations of single-sample peak callers in broad data \citep{stark2011diffbind,chen2015novel}. Moreover, most DPCs restrict their application to the analysis of two experimental conditions. For DPCs tailored for the analysis of three or more conditions, the classification of specific differential combinatorial patterns across conditions (or across various epigenomic processes) is still an open problem. The classification of such patterns would allow researchers to quantify treatment responses on the epigenomic level or identify sets of processes working together to regulate local chromatin state \citep{clouaire2014cfp1}.

@@

Here, $\mb{x}$ is a fixed $G \times L$ design matrix enumerating each of the $L$ possible differential combinatorial patterns in terms of the presence or absence of enrichment across each of the $G$ conditions, only in the emission distribution of the differential state. We assume that read counts pertaining to genomic windows from the consensus background ($r=1$) and consensus enrichment  ($r=3$) states follow a Negative Binomial (NB) distribution with state-specific parameters $\bs{\psi}_{r}=(\mu_{(r,hij)},\phi_{r})^\prime$, with mean $\mu_{(r,hij)}$ and variance $\mu_{(r,hij)}(1+\mu_{(r,hij)}/\phi_{r})$. Conditional on the HMM state, we assume independence of read counts across samples and write the emission distribution of the consensus background and enrichment states as
\begin{align}\label{eq:nondiffdensity}
f_{r}(\mb{y}_{..j}|\bs{\psi}_{r}) &= \prod^{G}_{h=1}\prod^{n_{h}}_{i=1}\frac{\Gamma(y_{hij}+\phi_{r})}{y_{hij}!\Gamma(\phi_{r})}\left(\frac{\phi_{r}}{\mu_{(r,hij)}+\phi_{r}}\right)^{\phi_{r}}\left(\frac{\mu_{(r,hij)}}{\mu_{(r,hij)}+\phi_{r}}\right)^{y_{hij}},\quad r\in\{1,3\},
\end{align}
such that $\log(\mu_{(1,hij)})=\beta_{1} + u_{hij}$, $\log(\phi_{1})=\lambda_{1}$, $\log(\mu_{(3,hij)})=\beta_{1} + \beta_{3} + u_{hij}$, and $\log(\phi_{3})=\lambda_{1} + \lambda_{3}$. The offset $u_{hij}$ may adjust for technical or biological artifacts such as the GC-content bias and allows the non-linear normalization of read counts across genomic windows, conditions, and replicates (Web Appendix B1; \cite{teng2017accounting}). %When $u_{hij}=0$, $\beta_{1}$ and $\lambda_{1}$ represent the log-mean and log-dispersion, respectively, of read counts pertaining to consensus background state windows, whereas $\beta_{3}$ and $\lambda_{3}$ represent the difference in log-mean and log-dispersion of read counts from consensus enrichment state windows relative to consensus background state windows.

In the differential state ($r = 2$), read counts are modeled by a $L$-component finite mixture model with mixture components that follow a Negative Binomial distribution, where each component corresponds to a particular differential combinatorial pattern. To define these patterns, consider sets $S_{1},\ldots,S_{L}$  delineating the subset of the $G$ conditions that are enriched in each of the $L$ differential combinatorial patterns. For instance, if $G=3$, the sets $S_{1} =\{1\}$, $S_{2} =\{2\}$, $S_{3} =\{3\}$, $S_{4} =\{1,2\}$, $S_{5} =\{1,3\}$, and $S_{6} =\{2,3\}$ define the six possible differential combinatorial patterns of enrichment and background across three conditions (\textit{e.g.}, $S_{6}$ denotes enrichment in conditions 2 and 3 and background in condition 1). The presence or absence of enrichment in each of the $L$ sets is encoded into each column of $\mb{x}=(\mb{x}_{1},\ldots,\mb{x}_{L})$, such that $\mb{x}_{l} = (x_{1l},\ldots,x_{Gl})'$, and $x_{hl} = I(h\in S_{l})$ for $l=1,\ldots,L$ and $h = 1,\ldots,G$. That is, $\mb{x}_{l}$ is the $G \times 1$ vector of binary indicator variables denoting which subset of conditions are enriched in pattern (mixture component) $l$. Let $\bs{\psi}_{2}$ and $\bs{\psi}_{(2,l)}$ denote the parameter vectors pertaining to the differential state and to the $l^{th}$ mixture component, respectively. Conditional on the differential state, we assume independence of read counts across samples and write the finite mixture model emission distribution as
\begin{align}\label{eq:diffdensity}
\resizebox{.925\hsize}{!}{$f_{2}(\mb{y}_{..j}|\mb{x};\bs{\delta},\bs{\psi}_{2}) = \sum_{l=1}^{L}\delta_{l}\prod_{h=1}^{G}\prod_{i=1}^{n_{h}}\frac{\Gamma\left(y_{hij}+\phi_{(2,lh)}\right)}{y_{hij}!\Gamma\left(\phi_{(2,lh)}\right)}\left(\frac{\phi_{(2,lh)}}{\mu_{(2,lhij)}+\phi_{(2,lh)}}\right)^{\phi_{(2,lh)}}\left(\frac{\mu_{(2,lhij)}}{\mu_{(2,lhij)}+\phi_{(2,lh)}}\right)^{y_{hij}},$}
\end{align}
where $\mu_{(2,lhij)}$ and $\phi_{(2,lh)}$ are the mean and dispersion, respectively, pertaining to read counts originating from window $j$ and sample $i$ in condition $h$ from the mixture component $l$. We assume that $\log(\mu_{(2,lhij)}) = \beta_{1} + \beta_{3}x_{hl} + u_{hij}$ and $\log(\phi_{(2,lh)}) = \lambda_{1} + \lambda_{3}x_{hl}$. That is, in the mixture component $l$, we utilize the same consensus background (consensus enriched) parametrization from (\ref{eq:nondiffdensity}) in all conditions that $\mb{x}_{l}$ specifies to be background (enriched) in the $l^{th}$ differential combinatorial pattern. Such a parametrization ensures that windows exhibiting differential enrichment across conditions share common means and dispersions between the consensus background enrichment states, an assumption that increases computational efficiency. Utilizing a mixture model as the differential state emission distribution avoids the computational burden that would come from assuming separate hidden states for each of the $L$ differential combinatorial patterns.  We evaluate the strength of these assumptions through multiple simulations and a real data benchmarking analysis in Sections \ref{sec:simulation} and \ref{sec:realdata}.

@@

The EM algorithm iterates until the maximum absolute relative change in the parameter estimates three iterations apart is less than $10^{-3}$ for three consecutive iterations. Here, we use a RCEM algorithm with threshold 0.05, which substantially reduces the dimensionality of the data during the M-step by randomly assigning a zero posterior probability to genomic windows unlikely to belong to each of the HMM states. Moreover, our estimation scheme allows genomic windows with equal distribution of counts across replicates and conditions to have their posterior probability aggregated during the M-step of the algorithm, which leads to a fast gradient-based optimization. Upon convergence, HMM posterior probabilities can be used to segment the genome into consensus background, differential, or consensus enrichment windows. Approaches that control the total false discovery rate (FDR) via posterior probabilities \citep{efron2001empirical} or that estimate the most likely sequence of hidden states can be used. Let $\hat{\rho}_{j2} = Pr\left(Z_{j}=2|\mb{y},\mb{x};\hat{\bs{\Psi}}\right)$ denote the estimated posterior probability that the $j^{th}$ genomic window belongs to the differential HMM state, $j=1,\ldots,M$. A posterior probability cutoff $\alpha$ is then chosen by controlling the total FDR $\sum_{j=1}^{M}(1-\hat{\rho}_{j2})I(\hat{\rho}_{j2}\geq 1-\alpha)/\sum_{j=1}^{M}I(\hat{\rho}_{j2}\geq 1-\alpha)$, where $I(\cdot)$ is an indicator function. Differential peaks are formed by merging adjacent windows that either meet a nominal FDR level for the differential HMM state or belong to the same predicted state (Web Appendix B3). Under this set up, we are able not only to detect differential enrichment regions across multiple conditions, but we can also classify various differential combinatorial patterns of enrichment within broad and short domains with mixture model posterior proportions.

The estimation scheme is robust to situations where certain differential combinatorial patterns of enrichment are rare (Figure \ref{fig:fig_multimarks}), which results from the fact that ChIP-seq experiments often provide enough data to estimate the parameters $(\beta_{1},\beta_{3},\lambda_{1},\lambda_{3})^\prime$ shared across all $L$ mixture components and HMM states. If pruning differential combinatorial patterns of the mixture model is of interest, the optimal number of mixture components $L^{*}$, $L^{*}<L$, can be selected via the Bayesian Information Criterion (BIC) for HMMs. We observed that selecting the optimal $L^{*}$ and combinatorial patterns via BIC agrees with pruning of rare differential combinatorial patterns that one would not expect to observe biologically (Web Appendix B2). In addition, the current implementation of epigraHMM allows the adjustment for external covariates, such as GC-content and input controls, via model offsets (Web Appendix A3). In real data analyses, however, we did not observe a significant influence of input controls on the sensitivity and specificity of peak calls, a fact that has also been noted by others \citep{lun2015csaw}.

@@

\subsection{Read Count Simulation}
\label{sec:sim1}
Read counts were simulated under scenarios that varied regarding the type of histone modification mark (H3K36me3 and H3K27me3), number of windows ($M$, $10^5$, $5\times 10^5$, and $10^6$ windows), number of conditions ($G$, 2, 3, and 4), and number of replicates per condition ($n$, 1, 2, and 4). We further assessed epigraHMM under different Signal-to-Noise Ratio (SNR) levels, here defined as the ratio between the means of consensus enrichment and background emission distributions, by decreasing the mean ratio while maintaining the mean-variance relationship. Parameters used in this study were estimated from ENCODE data (Web Appendix C1). Simulated counts followed a NB distribution and were generated using a first-order Markov chain with $2^G$ states, representing every combination of background and enrichment across $G$ conditions. We assessed whether epigraHMM was able to assign all $2^G-2$ differential states to the differential HMM state, while precisely estimating model parameters and accurately classifying differential combinatorial patterns.

@@

\subsection{Sequencing Read Simulation}
\label{sec:sim2}

Here, we compared epigraHMM with the DPCs ChIPComp, csaw, DiffBind, diffReps, RSEG, and THOR on data simulated with the pipeline from \cite{lun2015csaw} without a particular model assumption. Sequencing reads from ChIP-seq experiments were generated for two conditions and two replicates per condition. For the two-step DPCs ChIPComp and DiffBind, we followed \cite{lun2015csaw} and called peaks in advance using HOMER \citep{heinz2010simple}, which were then used as input in the respective software for differential call (MACS2 failed to output peaks on the simulated data). A hundred simulated data sets were generated and peaks were called under multiple nominal FDR thresholds. For epigraHMM and RSEG, window-based posterior probabilities were used to control the total FDR (Section \ref{subsec:hmmest}).

Figure~\ref{fig:fig_simulation_csaw} shows the results of our second simulation study. epigraHMM showed the highest observed sensitivity among all DPCs while maintaining an observed FDR close to the nominal value (Figure~\ref{fig:fig_simulation_csaw}A). Here, the observed FDR is the proportion of windows incorrectly called as differential out of the total number of windows called as differential. Due to the excessive number of false positive differential peak calls, diffReps, RSEG, and THOR showed higher observed FDR levels than the nominal FDR values (Figure~\ref{fig:fig_simulation_csaw}D). Overall, diffReps and THOR called an excessive number of short and discontiguous peaks while RSEG called regions that were usually wider than the simulated differential peaks (Figure~\ref{fig:fig_simulation_csaw}B). Regarding the computation time, epigraHMM was among the fastest DPCs due to the efficient RCEM implementation (Figure~\ref{fig:fig_simulation_csaw}C). Other HMM-based algorithms, RSEG and THOR, appeared to be the most computationally intensive and required longer amounts of time to analyze the data. In general, epigraHMM was able to consistently cover most of true differential regions with broad peaks while exhibiting a limited number of false discoveries (Figure~\ref{fig:fig_simulation_csaw}D; Web Appendix C2). Results shown for RSEG do not include 27 cases in which the algorithm failed to analyze the data due to internal errors or called the entire genome as differential. Similar issues of RSEG have been reported by others \citep{starmer2016detecting}.

@@

We compared the genome-wide performance of the presented model with the widely used DPCs ChIPComp, csaw, DiffBind, diffReps, RSEG, and THOR, a list that covers a variety of algorithms for both broad and short ChIP-seq enrichment profiles. For two-step DPCs, ChIPComp and DiffBind, we used peak calls from MACS2 as candidate peaks \citep{zhang2008model}. Methods were compared in terms of the coverage of differentially enriched genes, the number and average size of differential peak calls, the $\log_{2}$ fold change (LFC) and Spearman correlation of $\log_{2}$ reads counts mapped onto differential peak calls, the computation time, and maximum memory utilized. Metrics for sensitivity and specificity were defined on the window level and based on the coverage of differentially enriched genes by called peaks (Web Appendix D1). Read counts were computed using non-overlapping windows of 500bp and 250bp for broad and short marks, respectively (see Web Appendix D2 for a discussion on window size). Results presented here pertain to the analysis of cell lines Helas3 and Hepg2 (see Web Appendix D3 for additional results). Analyses were adjusted by input control samples in methods that are designed to do so (ChIPComp, DiffBind, diffReps, and THOR). We did not observe a significant improvement in performance by accounting for input control effect or GC-content bias with epigraHMM and did not adjust our analyses for these effects (Web Appendix B1).

\subsection{Analysis of ChIP-seq Data From Broad Marks}
\label{sec:broad}
First, we compared methods regarding the coverage of differentially enriched genes by differential peak calls for H3K36me3. Following the analysis of \cite{ji2013differential}, we defined a set of protein coding genes exhibiting $|\text{LFC}|>2$ of ChIP-seq counts between Helas3 and Hepg2 cell lines as true differential binding sites (see Web Appendix D1 for results using different thresholding values). Median ratio normalization of counts was performed on protein-coding genes after excluding those with total counts under the $25^{th}$ percentile of the distribution. Figure~\ref{fig:fig_broad}A shows receiver operating characteristic (ROC) curves for various methods and different nominal FDR levels for H3K36me3 differential peak calls. Observed true (false) positive rates were computed on the window-level as the proportion of windows called as differential out of all windows associated (not associated) with differentially enriched genes (Web Appendix D1). Overall, epigraHMM outperformed other DPCs by covering most of the differentially enriched genes while calling a limited number of false positive peaks.

>>>>

% Paragraph 3: Introduce ChIP-seq
To quantify local epigenomic activity, a common high-throughput assay is chromatin immunoprecipitation followed by massively parallel sequencing (ChIP-seq). ChIP-seq begins with cross-linking DNA and proteins within chromatin structures, which are then fragmented by sonication. DNA fragments bound to the protein of interest are isolated by immunoprecipitation and sequenced via high-throughput sequencing to generate short reads pertaining to the original fragments. Sequences are then mapped onto a reference genome to determine their likely locations of origin. Genomic coordinates containing a high density of mapped reads (enrichment regions, peaks) indicate likely locations of protein-DNA interaction sites, while other regions are referred to as background regions. The read density is often summarized by counting reads mapped onto non-overlapping windows of fixed length (window read counts), forming the basis for downstream analyses. Across multiple conditions, regions exhibiting enrichment in at least one condition, but not across all conditions, indicate the presence of differential activity pertaining to the protein-DNA interaction of interest.

@@

% Paragraph 4: What we propose? Broad peaks, mixture model, and multiple conditions
Here, we present epigraHMM (Figure~\ref{fig:fig_motivation}), an efficient and flexible statistical method to identify differential regions of enrichment from epigenomic experiments with diverse signal profiles. We overcome the limitations of current DPCs with three major features. First, it uses a hidden Markov model (HMM) to account for the diversity in differential enrichment from broad and short ChIP-seq data collected under multi-replicate, multi-condition settings. Second, it models specific combinatorial patterns of enrichment via a finite mixture model emission distribution within the HMM differential state. Each mixture component pertains to a particular differential combinatorial pattern originated by the presence or absence of local enrichment across conditions, where a generalized linear model (GLM) encodes specific differential combinatorial patterns while accounting for sample- and window-specific normalizing offsets. Third, it simultaneously detects and classifies epigenomic changes under three or more conditions, a novelty not yet available in any other DPC algorithm. epigraHMM offers benefits over current HMM algorithms \citep{song2011identifying,allhoff2016differential} with an embedded GLM mixture model that allows the modeling of covariates of interest, the inclusion of normalizing offsets for non-linear biases (such as GC-content bias), and a fast estimation scheme via rejection-controlled EM algorithm (RCEM; \cite{ma2006data}).

@@

Using ChIP-seq data pertaining to histone modifications H3K27me3 and H3K36me3, and to EZH2 from the ENCODE Consortium \citep{landt2012chip}, we find that current DPCs have difficulty in accurately detecting broad regions of differential enrichment between several common cell lines (Figure~\ref{fig:fig_motivation}). In line with previous findings \citep{steinhauser2016comprehensive}, we observe that even current DPCs designed for broad data \citep{song2011identifying,allhoff2016differential} tend to either call overly fragmented differential peaks or call regions exhibiting no difference in experimental signal between conditions as differential (Sections \ref{sec:sim2} and \ref{sec:broad}). Methods that rely on candidate peaks may also exhibit a compromised performance due to the limitations of single-sample peak callers in broad data \citep{stark2011diffbind,chen2015novel}. Moreover, most DPCs restrict their application to the analysis of two experimental conditions. For DPCs tailored for the analysis of three or more conditions, the classification of specific differential combinatorial patterns across conditions (or across various epigenomic processes) is still an open problem. The classification of such patterns would allow researchers to quantify treatment responses on the epigenomic level or identify sets of processes working together to regulate local chromatin state \citep{clouaire2014cfp1}.

@@

Here, $\mb{x}$ is a fixed $G \times L$ design matrix enumerating each of the $L$ possible differential combinatorial patterns in terms of the presence or absence of enrichment across each of the $G$ conditions, only in the emission distribution of the differential state. We assume that read counts pertaining to genomic windows from the consensus background ($r=1$) and consensus enrichment  ($r=3$) states follow a Negative Binomial (NB) distribution with state-specific parameters $\bs{\psi}_{r}=(\mu_{(r,hij)},\phi_{r})^\prime$, with mean $\mu_{(r,hij)}$ and variance $\mu_{(r,hij)}(1+\mu_{(r,hij)}/\phi_{r})$. Conditional on the HMM state, we assume independence of read counts across samples and write the emission distribution of the consensus background and enrichment states as
\begin{align}\label{eq:nondiffdensity}
f_{r}(\mb{y}_{..j}|\bs{\psi}_{r}) &= \prod^{G}_{h=1}\prod^{n_{h}}_{i=1}\frac{\Gamma(y_{hij}+\phi_{r})}{y_{hij}!\Gamma(\phi_{r})}\left(\frac{\phi_{r}}{\mu_{(r,hij)}+\phi_{r}}\right)^{\phi_{r}}\left(\frac{\mu_{(r,hij)}}{\mu_{(r,hij)}+\phi_{r}}\right)^{y_{hij}},\quad r\in\{1,3\},
\end{align}
such that $\log(\mu_{(1,hij)})=\beta_{1} + u_{hij}$, $\log(\phi_{1})=\lambda_{1}$, $\log(\mu_{(3,hij)})=\beta_{1} + \beta_{3} + u_{hij}$, and $\log(\phi_{3})=\lambda_{1} + \lambda_{3}$. The offset $u_{hij}$ may adjust for technical or biological artifacts such as the GC-content bias and allows the non-linear normalization of read counts across genomic windows, replicates, and conditions (Web Appendix B1; \cite{teng2017accounting}). %When $u_{hij}=0$, $\beta_{1}$ and $\lambda_{1}$ represent the log-mean and log-dispersion, respectively, of read counts pertaining to consensus background state windows, whereas $\beta_{3}$ and $\lambda_{3}$ represent the difference in log-mean and log-dispersion of read counts from consensus enrichment state windows relative to consensus background state windows.

In the differential state ($r = 2$), read counts are modeled by a $L$-component finite mixture model with mixture components that follow a NB distribution, where each component corresponds to a particular differential combinatorial pattern. To define these patterns, consider sets $S_{1},\ldots,S_{L}$  delineating the subset of the $G$ conditions that are enriched in each of the $L$ differential combinatorial patterns. For instance, if $G=3$, the sets $S_{1} =\{1\}$, $S_{2} =\{2\}$, $S_{3} =\{3\}$, $S_{4} =\{1,2\}$, $S_{5} =\{1,3\}$, and $S_{6} =\{2,3\}$ define the six possible differential combinatorial patterns of enrichment and background across three conditions (\textit{e.g.}, $S_{6}$ denotes enrichment in conditions 2 and 3 and background in condition 1). The presence or absence of enrichment in each of the $L$ sets is encoded into each column of $\mb{x}=(\mb{x}_{1},\ldots,\mb{x}_{L})$, such that $\mb{x}_{l} = (x_{1l},\ldots,x_{Gl})'$, and $x_{hl} = I(h\in S_{l})$ for $l=1,\ldots,L$ and $h = 1,\ldots,G$. That is, $\mb{x}_{l}$ is the $G \times 1$ vector of binary indicator variables denoting which subset of conditions are enriched in pattern (mixture component) $l$. Let $\bs{\psi}_{2}$ and $\bs{\psi}_{(2,l)}$ denote the parameter vectors pertaining to the differential state and to the $l^{th}$ mixture component, respectively. Conditional on the differential state, we assume independence of read counts across samples and write the finite mixture model emission distribution as
\begin{align}\label{eq:diffdensity}
\resizebox{.925\hsize}{!}{$f_{2}(\mb{y}_{..j}|\mb{x};\bs{\delta},\bs{\psi}_{2}) = \sum_{l=1}^{L}\delta_{l}\prod_{h=1}^{G}\prod_{i=1}^{n_{h}}\frac{\Gamma\left(y_{hij}+\phi_{(2,lh)}\right)}{y_{hij}!\Gamma\left(\phi_{(2,lh)}\right)}\left(\frac{\phi_{(2,lh)}}{\mu_{(2,lhij)}+\phi_{(2,lh)}}\right)^{\phi_{(2,lh)}}\left(\frac{\mu_{(2,lhij)}}{\mu_{(2,lhij)}+\phi_{(2,lh)}}\right)^{y_{hij}},$}
\end{align}
where $\mu_{(2,lhij)}$ and $\phi_{(2,lh)}$ are the mean and dispersion, respectively, pertaining to read counts originating from window $j$ and sample $i$ in condition $h$ from the mixture component $l$. We assume that $\log(\mu_{(2,lhij)}) = \beta_{1} + \beta_{3}x_{hl} + u_{hij}$ and $\log(\phi_{(2,lh)}) = \lambda_{1} + \lambda_{3}x_{hl}$. That is, in the mixture component $l$, we utilize the same consensus background (consensus enriched) parametrization from (\ref{eq:nondiffdensity}) in all conditions that $\mb{x}_{l}$ specifies to be background (enriched) in the $l^{th}$ differential combinatorial pattern. Such a parametrization ensures that windows exhibiting differential enrichment across conditions share common means and dispersions between the consensus background enrichment states, an assumption that increases computational efficiency. Utilizing a mixture model as the differential state emission distribution avoids the computational burden that would come from assuming separate hidden states for each of the $L$ differential combinatorial patterns.  We evaluate the strength of these assumptions through multiple simulations and a real data benchmarking analysis in Sections \ref{sec:simulation} and \ref{sec:realdata}.

@@

The EM algorithm iterates until the maximum absolute relative change in the parameter estimates three iterations apart is less than $10^{-3}$ for three consecutive iterations. Here, we use a RCEM algorithm with threshold 0.05, which substantially reduces the dimensionality of the data during the M-step by randomly assigning a zero posterior probability to genomic windows unlikely to belong to each of the HMM states. Moreover, our estimation scheme allows genomic windows with equal distribution of counts across replicates and conditions to have their posterior probability aggregated during the M-step of the algorithm, which leads to a fast gradient-based optimization. Upon convergence, HMM posterior probabilities can be used to segment the genome into consensus background, differential, or consensus enrichment windows. Approaches that control the total false discovery rate (FDR) via posterior probabilities \citep{efron2001empirical} or that estimate the most likely sequence of hidden states can be used. Let $\hat{\rho}_{j2} = Pr\left(Z_{j}=2|\mb{y},\mb{x};\hat{\bs{\Psi}}\right)$ denote the estimated posterior probability that the $j^{th}$ genomic window belongs to the differential HMM state, $j=1,\ldots,M$. A posterior probability cutoff $\alpha$ is then chosen by controlling the total FDR $\sum_{j=1}^{M}(1-\hat{\rho}_{j2})I(\hat{\rho}_{j2}\geq 1-\alpha)/\sum_{j=1}^{M}I(\hat{\rho}_{j2}\geq 1-\alpha)$, where $I(\cdot)$ is an indicator function. Differential peaks are formed by merging adjacent windows that either meet a nominal FDR level for the differential HMM state or belong to the same predicted state (Web Appendix B3). Under this set up, we are able not only to detect differential enrichment regions across multiple conditions, but we can also classify various differential combinatorial patterns of enrichment within broad and short domains with mixture model posterior probabilities.

The estimation scheme is robust to situations where certain differential combinatorial patterns of enrichment are rare (Figure \ref{fig:fig_multimarks}), which results from the fact that ChIP-seq experiments often provide enough data to estimate the parameters $(\beta_{1},\beta_{3},\lambda_{1},\lambda_{3})^\prime$ shared across all $L$ mixture components and HMM states. If pruning differential combinatorial patterns of the mixture model is of interest, the optimal number of mixture components $L^{*}$, $L^{*}<L$, can be selected via the Bayesian Information Criterion (BIC) for HMMs. We observed that selecting the optimal $L^{*}$ and combinatorial patterns via BIC agrees with pruning of rare differential combinatorial patterns that one would not expect to observe biologically (Web Appendix B2). In addition, the current implementation of epigraHMM allows for the adjustment for external covariates, such as GC-content and input controls, via model offsets (Web Appendix A3). In real data analyses, however, we did not observe a significant influence of input controls on the sensitivity and specificity of differential peak calls, a fact that has also been noted by others \citep{lun2015csaw}.

@@

\subsection{Read Count Simulation}
\label{sec:sim1}
Read counts were simulated under scenarios that varied regarding the type of histone modification mark (H3K36me3 and H3K27me3), number of windows ($M$, $10^5$, $5\times 10^5$, and $10^6$ windows), number of conditions ($G$, 2, 3, and 4), and number of replicates per condition ($n$, 1, 2, and 4). We further assessed epigraHMM under different Signal-to-Noise Ratio (SNR) levels, here defined as the ratio between the means of consensus enrichment and background emission distributions, by decreasing the mean ratio while maintaining the mean-variance relationship. Model parameters were estimated from ENCODE data (Web Appendix C1). Simulated counts followed a NB distribution and were generated using a first-order Markov chain with $2^G$ states, representing every combination of background and enrichment across $G$ conditions. We assessed whether epigraHMM was able to assign all $2^G-2$ differential states to the differential HMM state, while precisely estimating model parameters and accurately classifying differential combinatorial patterns.

@@

\subsection{Sequencing Read Simulation}
\label{sec:sim2}
Here, we compared epigraHMM with the widely used DPCs ChIPComp, csaw, DiffBind, diffReps, RSEG, and THOR, a list that covers a variety of algorithms for both broad and short ChIP-seq enrichment profiles, on data simulated with the pipeline from \cite{lun2015csaw} without a particular model assumption. Sequencing reads from ChIP-seq experiments were generated for two conditions and two replicates per condition. For the two-step DPCs ChIPComp and DiffBind, we followed \cite{lun2015csaw} and called peaks in advance using HOMER \citep{heinz2010simple}, which were then used as input in the respective software for differential call. A hundred simulated data sets were generated and peaks were called under multiple nominal FDR thresholds. For epigraHMM and RSEG, window-based posterior probabilities were used to control the total FDR (Section \ref{subsec:hmmest}).

Overall, epigraHMM showed the highest observed sensitivity among all DPCs while maintaining an observed FDR close to the nominal value across a wide range of nominal FDR levels (Figure~\ref{fig:fig_simulation_csaw}A). Here, the observed FDR is the proportion of windows incorrectly called as differential out of the total number of windows called as differential. Due to the excessive number of false positive differential peak calls, diffReps, RSEG, and THOR showed higher observed FDR levels than the nominal FDR values (Figure~\ref{fig:fig_simulation_csaw}D). Overall, diffReps and THOR called an excessive number of short and discontiguous peaks while RSEG called regions that were usually wider than the simulated differential peaks and did not correspond to simulated differential enrichment (Figure~\ref{fig:fig_simulation_csaw}B). Regarding the computation time, epigraHMM was among the fastest DPCs due to the efficient RCEM implementation (Figure~\ref{fig:fig_simulation_csaw}C). Other HMM-based algorithms, RSEG and THOR, appeared to be the most computationally intensive and required longer amounts of time to analyze the data. In general, epigraHMM was able to consistently cover most of true differential regions with broad peaks while exhibiting a limited number of false discoveries (Figure~\ref{fig:fig_simulation_csaw}D; Web Appendix C2). Results shown for RSEG do not include 27 cases in which the algorithm failed to analyze the data due to internal errors or called the entire genome as differential. Similar issues of RSEG have been reported by others \citep{starmer2016detecting}.

@@

We compared the genome-wide performance of epigraHMM with ChIPComp, csaw, DiffBind, diffReps, RSEG, and THOR. For two-step DPCs, ChIPComp and DiffBind, we used peak calls from MACS2 as candidate peaks \citep{zhang2008model}. Methods were compared in terms of the coverage of differentially enriched genes, the number and average size of differential peak calls, the $\log_{2}$ fold change (LFC) and Spearman correlation of $\log_{2}$ reads counts mapped onto differential peak calls, the computation time, and maximum memory utilized. Metrics for sensitivity and specificity were defined on the window level and based on the coverage of differentially enriched genes by called peaks (Web Appendix D1). Read counts were computed using non-overlapping windows of 500bp and 250bp for broad and short marks, respectively (see Web Appendix D2 for a discussion on window size). Results presented here pertain to the analysis of cell lines Helas3 and Hepg2 (see Web Appendix D3 for additional results). Analyses were adjusted by input control samples in methods that are designed to do so (ChIPComp, DiffBind, diffReps, and THOR). We did not observe a significant improvement in performance by accounting for input control effect or GC-content bias with epigraHMM and did not adjust our analyses for these effects (Web Appendix B1).

\subsection{Analysis of ChIP-seq Data From Broad Marks}
\label{sec:broad}
We compared methods regarding the coverage of differentially enriched genes by differential peak calls for H3K36me3, an epigenomic mark associated with transcribed genes. Following \cite{ji2013differential}, in which the authors define true differential binding sites from LFCs of ChIP-seq counts to assess model sensitivity (Figure 2E in \cite{ji2013differential}), we defined a set of protein coding genes exhibiting $|\text{LFC}|>2$ of ChIP-seq counts between Helas3 and Hepg2 cell lines as true differential binding sites (see Web Appendix D1 for results using different thresholding values). Median ratio normalization of counts was performed on protein-coding genes after excluding those with total counts under the $25^{th}$ percentile of the distribution. Figure~\ref{fig:fig_broad}A shows receiver operating characteristic (ROC) curves for various methods and different nominal FDR levels for H3K36me3 differential peak calls. Observed true (false) positive rates were computed on the window-level as the proportion of windows called as differential out of all windows associated (not associated) with differentially enriched genes (Web Appendix D1). Overall, epigraHMM outperformed other DPCs by covering most of the differentially enriched genes while calling a limited number of false positive peaks.

DiffBind, which has been shown to be dependent on the set of candidate peaks \citep{lun2015csaw}, called the shortest peaks overall (Figures \ref{fig:fig_broad}B and \ref{fig:fig_broad}C) and ChIPComp, which has been shown to perform best in scenarios with short marks \citep{steinhauser2016comprehensive}, appeared to have limited performance in broad marks such as H3K27me3 and EZH2 (Figures \ref{fig:fig_broad}D and \ref{fig:fig_broad}F). epigraHMM and RSEG, two HMM-based methods, called broader differential peaks and exhibited better sensitivity than other methods. Yet, differential peaks from RSEG often did not correspond to observed differential enrichment, a fact that explains its high observed FDR and agrees with our simulations (Figures \ref{fig:fig_broad}C and \ref{fig:fig_simulation_csaw}) and others \citep{starmer2016detecting}. Our HMM-based approach with a non-linear normalization via model offsets allowed us to maintain a low observed FDR and a higher sensitivity than other DPCs. Overall, epigraHMM was among the most efficient algorithms due to our computational scheme, taking approximately 30 minutes to analyze genome-wide data (Figure~\ref{fig:fig_broad}E).
